Learning Theory Notes

### 2022-06-05 notebooks/theory_notes.tex
updated notes with error model

>>>>

\section*{Multiplicative Error Model}

The predicted reduction for segment $i$ from the machine learned model
$\hat{B}_i$ can be partitioned as $\hat{B}_i = B_i + \epsilon_i$, where
$\epsilon_i$ is the cumulative error due to bias, irreducible evolutionary
variance, and training variance. Then, our total predicted reduction at some
focal site $v$ is:

\begin{align}
  \hat{B}(v) = \prod_i^n (B_i + \epsilon_i)
\end{align}

We want to understand the bias and variance of this product estimator based on
the total learning error. Taking logs,

\begin{align}
  \hat{B}(v) &= \prod_i^n (B_i + \epsilon_i) \\
  \log(\hat{B}(v)) &= \log\left(\prod_i^n (B_i + \epsilon_i) \right) \\
                   &= \sum_i^n \log(B_i + \epsilon_i). \\
\end{align}

If we approximate the expected log with a Taylor series,

\begin{align}
  \E[\log(B_i + \epsilon_i)^2] &\approx \log(B_i)^2 + \frac{2 \log(B_i)}{B_i} \E[\epsilon_i] + \frac{(1-\log(B_i))}{B_i^2} \E[\epsilon_i^2]\\
                               &\approx \log(B_i)^2 + \frac{2 \log(B_i)}{B_i}b + \frac{(1-\log(B_i)) }{B_i^2}(\sigma^2 + b^2)
\end{align}

Then,

\begin{align}
  \E[\log(\hat{B}(v))] &= \sum_i^n \E[\log(B_i + \epsilon_i)]  \\
                                        &\approx \sum_i^n \log(B_i) + \sum_i^n \frac{b}{B_i} - \sum_i^n \frac{\sigma^2 + b^2}{2B_i^2}  \\
\end{align}

Bias:

\begin{align}
  \E[\log(\hat{B}(v))] - \E[\log(B(v))] &= \sum_i^n \E[\log(B_i + \epsilon_i)] - \sum_i^n \log(B_i) \\
                                        &\approx \sum_i^n \log(B_i) + \sum_i^n \frac{b}{B_i} - \sum_i^n \frac{\sigma^2 + b^2}{2B_i^2} - \sum_i^n \log(B_i) \\
                                        &\approx \sum_i^n \frac{b}{B_i} - \sum_i^n \frac{\sigma^2 + b^2}{2B_i^2} 
\end{align}

By Jensen's inequality $f(\E(X)) \le \E(f(X))$ for some convex function $f$,

\section*{Likelihood}

The likelihood in Eyalshiv et al. (2016) has the form:

\begin{align}
  \log\mathcal{L} = \sum_{v \in V} \sum_{i \ne j \in S} \log(P(O_{i,j}(v) | \theta))
\end{align}

where $V$ is the collection of neutral sites, $S$ is the set of samples, and
$\theta$ are the BGS parameters. The indicator variable $O_{i,j}(v)$ is 1 if
samples $i$ and $j$ are different at site $v$, and zero otherwise. Thus as they
specify in the paper,

\begin{align}
  P(O_{i,j}(v) | \theta) = 
    \begin{cases}
      \pi(v | \theta), & O_{i,j}(v) = 1 \\
      1-\pi(v | \theta), & O_{i,j}(v) = 0 \\
    \end{cases}.
\end{align}

For $S$ samples, there are $|S| \choose 2$ pairwise comparisons that we can
partition into the different and same classes; we assign their counts to
$n_\text{D}(v)$ and $n_\text{S}(v)$. Then,

\begin{align}
  \log\mathcal{L} = \sum_{v \in V} \left(\log(\pi(v | \theta)) n_\text{S}(v) + \log(1-\pi(v | \theta)) n_\text{D}(v)\right)
\end{align}

Since $B$ is discretized into genomic windows of length $K$ basepairs, $\pi(v |
\theta)$ is constant in a window.
